---
title: "Initial Descriptive Analysis of ECLS-K Achievement Data"
author: "SCALES Lab"
date: "2026-05-15"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This notebook provides an initial descriptive analysis of the ECLS-K achievement file used in the SVR-DD project. The goal is to inspect the structure of the dataset, summarize missingness, describe broad developmental patterns across grades, and examine within-grade associations among achievement domains before moving to formal developmental models.

## Setup

```{r}
library(tidyverse)
library(janitor)
library(knitr)
```

```{r}
eclsk <- read_csv(
  "../../data/eclsk_data.csv",
  na = c("", "NA")
) %>%
  clean_names() %>%
  select(-any_of(c("x", "x1")))

glimpse(eclsk)
```

## Data overview

The dataset contains repeated achievement measures for science, reading, and mathematics in grades 3, 5, and 8. Variables with prefixes `s_`, `r_`, and `m_` are the raw achievement measures used in this notebook. Variables with prefixes `st_`, `rt_`, and `mt_` are transformed versions retained in the source file but not emphasized in this initial overview.

```{r}
overview <- tibble(
  students = nrow(eclsk),
  variables = ncol(eclsk),
  grades = "3, 5, 8",
  domains = "Science, Reading, Mathematics"
)

kable(overview)
```

```{r}
raw_long <- eclsk %>%
  select(id, matches("^[srm]_g[358]$")) %>%
  pivot_longer(
    cols = -id,
    names_to = c("domain", "grade"),
    names_pattern = "([srm])_g([358])",
    values_to = "score"
  ) %>%
  mutate(
    domain = recode(
      domain,
      s = "Science",
      r = "Reading",
      m = "Mathematics"
    ),
    grade = as.integer(grade)
  )
```

## Missing-data summary

Because the same students are not observed at every wave, the first descriptive question is how much usable data are available for each domain and grade.

```{r}
missing_summary <- raw_long %>%
  group_by(domain, grade) %>%
  summarize(
    n_nonmissing = sum(!is.na(score)),
    n_missing = sum(is.na(score)),
    pct_nonmissing = mean(!is.na(score)) * 100,
    .groups = "drop"
  ) %>%
  arrange(domain, grade)

kable(
  missing_summary,
  digits = 1,
  col.names = c("Domain", "Grade", "Non-missing n", "Missing n", "Non-missing %")
)
```

The number of observed scores declines across grades, which will matter for any later longitudinal analysis.

## Descriptive statistics

```{r}
descriptives <- raw_long %>%
  group_by(domain, grade) %>%
  summarize(
    n = sum(!is.na(score)),
    mean = mean(score, na.rm = TRUE),
    sd = sd(score, na.rm = TRUE),
    min = min(score, na.rm = TRUE),
    max = max(score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(domain, grade)

kable(
  descriptives,
  digits = 2,
  col.names = c("Domain", "Grade", "n", "Mean", "SD", "Min", "Max")
)
```

## Mean achievement across grades

```{r}
mean_plot_data <- raw_long %>%
  group_by(domain, grade) %>%
  summarize(
    mean_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(mean_plot_data, aes(x = grade, y = mean_score, color = domain)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = c(3, 5, 8)) +
  labs(
    title = "Mean achievement increases from grade 3 to grade 8",
    x = "Grade",
    y = "Mean score",
    color = "Domain"
  ) +
  theme_minimal(base_size = 12)
```

All three domains show higher mean scores at later grades. This figure is descriptive only; it does not distinguish developmental growth from differences introduced by attrition or measurement features.

## Within-grade associations among domains

A second useful descriptive question is whether science, reading, and mathematics scores are positively associated within each grade.

```{r}
correlation_tables <- raw_long %>%
  select(id, domain, grade, score) %>%
  pivot_wider(names_from = domain, values_from = score) %>%
  group_by(grade) %>%
  group_map(~ {
    cor_matrix <- .x %>%
      select(Science, Reading, Mathematics) %>%
      cor(use = "pairwise.complete.obs") %>%
      round(2)

    list(
      grade = unique(.x$grade),
      matrix = cor_matrix
    )
  })
```

```{r, results='asis'}
for (item in correlation_tables) {
  cat("### Grade ", item$grade, "\n\n", sep = "")
  print(kable(item$matrix))
  cat("\n\n")
}
```

```{r}
cor_plot_data <- raw_long %>%
  select(id, domain, grade, score) %>%
  pivot_wider(names_from = domain, values_from = score) %>%
  group_by(grade) %>%
  summarize(
    science_reading = cor(Science, Reading, use = "pairwise.complete.obs"),
    science_math = cor(Science, Mathematics, use = "pairwise.complete.obs"),
    reading_math = cor(Reading, Mathematics, use = "pairwise.complete.obs"),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -grade,
    names_to = "pair",
    values_to = "correlation"
  ) %>%
  mutate(
    pair = recode(
      pair,
      science_reading = "Science–Reading",
      science_math = "Science–Mathematics",
      reading_math = "Reading–Mathematics"
    )
  )

ggplot(cor_plot_data, aes(x = factor(grade), y = correlation, fill = pair)) +
  geom_col(position = "dodge") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Achievement domains are positively associated within grade",
    x = "Grade",
    y = "Correlation",
    fill = "Domain pair"
  ) +
  theme_minimal(base_size = 12)
```

The domains are positively associated within each grade. These associations are consistent with the idea that reading-related and broader academic skills develop together, but they do not by themselves identify the mechanism producing that pattern.

## Initial observations

This first descriptive pass suggests four practical points for later modeling work:

1. The file contains repeated measures for three achievement domains across grades 3, 5, and 8.
2. Available sample size declines across later grades, so longitudinal analyses will need an explicit missing-data strategy.
3. Mean achievement scores rise across grades in all three domains.
4. Science, reading, and mathematics scores are positively associated within grade.

These patterns provide useful orientation for subsequent SVR-DD analyses, but they remain descriptive. Model comparison will require explicit developmental mechanisms and tests of the patterns those mechanisms imply.

## Session information

```{r}
sessionInfo()
```
